Fusing Filters with Integer Linear Programming

### 2014-05-05 2014betterfusionforfilters/section/02-CNF.tex
cnf: add todo for filter vs general size changing operators
Edits to Section 2

Once we have determined how the operators in a data flow graph should be clustered, we highlight materialized array variables by drawing them in boxes. In Figure~\ref{f:normalize2-clusterings}, the variables @xs@, @ys1@ and @ys2@ are always in boxes as these are the material input and output arrays of the program. However, in the graph on the far right hand side @gts@ has also been materialized because in this version the producing and consuming operators (@filter@ and @fold@) have not been fused.

In Figure~\ref{f:CombinatorNormalForm} note that the bindings have been split into the ones that produce scalar values ($sbind$), and the ones that produce array values ($abind$), and these groupings are represented as open and closed arrow-heads in Figure~\ref{f:normalize2-clusterings}.

Most of our array combinators are standard, and suggestive types are given at the bottom of Figure~\ref{f:CombinatorNormalForm}. The $@map@_n$ combinator takes a worker function, $n$ arrays of the same length, and applies the worker function to all elements at the same index. As such it is similar to Haskell's @zipWith@, except for the length restriction on the argument arrays. The @generate@ combinator takes an array length and a worker function, and creates a new array by applying the worker to each index. The @gather@ combinator takes an array of elements, an array of indices, and produces the array of elements that were at each index. In Haskell this would be @gather arr ixs = map (index arr) ixs@. The @cross@ combinator returns the cartesian product of two arrays.

@@

The @external@ binding invokes a host library function that can produce and consume arrays, but not be fused with other combinators. All arrays passed to, and returned from host functions are full materialised. External bindings are explicit \emph{fusion barriers}, which force arrays and scalars to be fully computed before continuing.

% When the function computed is the identity, they can also be used to force a particular intermediate array to be materialized, which provides the programmer fine-grained control over exactly what clustering is produced by our algorithm.

>>>>

Clusters of operators that are fused into single imperative loops are indicated by dotted lines, and we highlight materialized arrays by drawing them in boxes. In Figure~\ref{f:normalize2-clusterings}, the variables @xs@, @ys1@ and @ys2@ are always in boxes as these are the material input and output arrays of the program. However, in the graph on the far right hand side @gts@ has also been materialized because in this version the producing and consuming operators (@filter@ and @fold@) have not been fused. In Figure~\ref{f:CombinatorNormalForm} note that the bindings have been split into the ones that produce scalar values ($sbind$), and the ones that produce array values ($abind$), and these groupings are represented as open and closed arrow-heads in Figure~\ref{f:normalize2-clusterings}.

Most of our array combinators are standard, and suggestive types are given at the bottom of Figure~\ref{f:CombinatorNormalForm}. The $@map@_n$ combinator takes a worker function, $n$ arrays of the same length, and applies the worker function to all elements at the same index. As such it is similar to Haskell's @zipWith@, with an added length restriction on the argument arrays. The @generate@ combinator takes an array length and a worker function, and creates a new array by applying the worker to each index. The @gather@ combinator takes an array of elements, an array of indices, and produces the array of elements that were at each index. In Haskell this would be @gather arr ixs = map (index arr) ixs@. The @cross@ combinator returns the cartesian product of two arrays.

@@

The @external@ binding invokes a host library function that can produce and consume arrays, but not be fused with other combinators. All arrays passed to, and returned from host functions are fully materialised. External bindings are explicit \emph{fusion barriers}, which force arrays and scalars to be fully computed before continuing.

Finally, note that @filter@ is only one representative size changing operator. We can handle more complex functions such as @unfold@ in our framework, but we stick with simple filtering to aid the discussion.
